---
title: 'Project 2: Data Mining, Classification, Prediction'
author: "SDS322E"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))

class_diag <- function(score, truth, positive, cutoff=.5){

  pred <- factor(score>cutoff,levels=c("TRUE","FALSE"))
  truth <- factor(truth==positive, levels=c("TRUE","FALSE"))

  tab<-table(truth, pred)
  acc=sum(diag(tab))/sum(tab)
  sens=tab[1,1]/rowSums(tab)[1]
  spec=tab[2,2]/rowSums(tab)[2]
  ppv=tab[1,1]/colSums(tab)[1]

#CALCULATE F1
  f1=2*(sens*ppv)/(sens+ppv)
  
#CALCULATE EXACT AUC
  truth<-as.numeric(truth=="TRUE")
  ord<-order(score, decreasing=TRUE)
  score <- score[ord]; truth <- truth[ord]
  TPR=cumsum(truth)/max(1,sum(truth))
  FPR=cumsum(!truth)/max(1,sum(!truth))
  dup<-c(score[-1]>=score[-length(score)], FALSE)
  TPR<-c(0,TPR[!dup],1); FPR<-c(0,FPR[!dup],1)
  n <- length(TPR)
  auc<- sum( ((TPR[-1]+TPR[-n])/2) * (FPR[-1]-FPR[-n]) )
  round(data.frame(acc,sens,spec,ppv,f1,ba=(sens+spec)/2,auc, row.names = "Metrics"),4)
}
```

# Mining, Classification, Prediction

## Anne Siri Snell, as223795

### Introduction 

My data is from the FiveThirtyEight data package already built into R. The specific dataset i the 'hatecrimes' dataset, which contains data on the number of hate crimes on the state level along with id variable for the state and 9 different demographic variables. The variables of my interest are number of hate crimes, income, share of non-white population, share of population living in metropolitan areas and a binary variable I've created that indicated whether or not the state is above or below the mean share of adult population with a high school degree.

Income is measured as annual median household income in 2016, share of non-white population is share of the population that is not white in 2015 and ranges from 0-1, number of hate crimes is measured as per 100,000 population in 2016, share of the population that lives in metropolitan areas is measured in 2015 and ranges from 0-1. These are all numeric variables. 

My binary variable 'highschool' is given a 0 if the state have a share of population with a high school degree below the mean and 1 if the state have a population with a high school degree above the mean. 

There are 765 observations in my dataset and for the binary variable i have 23 observations below the mean of share of population with a high school degree and 28 observations above the mean. 

```{R}
#Load packages 
library(tidyverse)
library(fivethirtyeight)
library(cluster)
library(dplyr)
library(ggplot2)

#Show dataframe
print(hate_crimes)

#Count number of observations
length(as.matrix(hate_crimes))

#Create new binary varible "highschool"
mean(hate_crimes$share_pop_hs)
hate_crimes$highschool <- ifelse(hate_crimes$share_pop_hs>0.8691176,1,0)

#Show number of observations in each category of 'highschool'
table(hate_crimes$highschool)
```

### Cluster Analysis

```{R}
#Choosing number of clusters
pam_dat <- hate_crimes %>% select(median_house_inc, share_non_white, share_pop_metro)
sil_width<-vector()
for(i in 2:10){  
  pam_fit <- pam(pam_dat, k = i)  
  sil_width[i] <- pam_fit$silinfo$avg.width  
}
ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)
```
From the above plot I suggest to choose three clusters. 

```{r}
#Running cluster analysis 
pam1 <- pam_dat %>% pam(k=3) #use the pam function
pam1
```
```{r}
#Saving clusters to dataframe
hate_crimes <- hate_crimes %>% mutate(cluster=as.factor(pam1$clustering))

#Visualizing clusters
library(GGally)
ggpairs(hate_crimes, columns = c("median_house_inc", "share_non_white", "share_pop_metro"), aes(color=cluster))
```
```{r}
#Calculating average silhouette width
pam1$silinfo$avg.width
```

As the average silhouette width is between 0.51−0.70 a reasonable structure have been found. 
    
### Dimensionality Reduction with PCA

```{R}
#PCA
hate_crimes_nums<-hate_crimes %>% select(median_house_inc, share_non_white, share_pop_metro) %>% scale
hate_crimes_pca<-princomp(hate_crimes_nums)
summary(hate_crimes_pca, loadings=T)

eigval<-hate_crimes_pca$sdev^2
varprop=round(eigval/sum(eigval), 2)
round(cumsum(eigval)/sum(eigval), 2) #cumulative proportion of variance
eigval #eigenvalues

ggplot() + geom_bar(aes(y=varprop, x=1:3), stat="identity") + xlab("") + geom_path(aes(y=varprop, x=1:3)) + 
  geom_text(aes(x=1:3, y=varprop, label=round(varprop, 2)), vjust=1, col="white", size=5) + 
  scale_y_continuous(breaks=seq(0, .6, .2), labels = scales::percent) + 
  scale_x_continuous(breaks=1:10)

```

There are several rules of thumb to chose the right number of PCs. One is to look at when the scree plot flattens, but it isn't really clear from my plot when it flattens. Another startegy is to pick the number of PCs until cumulative proportion of variance is >80%. This would make me chose 2 PCs. A third rule is to pick PCs whose eigenvalues are greater than 1 (Kaiser's rule). This rule woudl make me choose only the first PC. All combined I choose PC1 and PC2, which explains 86 % of the variance in my data. PC1 shows positive and relatively high loadings on both income, share of non-white population and share of population in metropolitan areas. This means that the higher score on PC1, the higher overall score on the included variables. PC2 shows that higher scores on PC2 mean high median household income but low score on share of non-white population. 

###  Linear Classifier

```{R}
#Replacing NAs with 0
hate_crimes[is.na(hate_crimes)] <- 0

#Logistic regression as linear classifier
fit <- glm(highschool ~ median_house_inc + share_non_white + share_pop_metro + hate_crimes_per_100k_splc, data=hate_crimes, family="binomial")

score <- predict(fit, type="response")

class_diag(score,truth=hate_crimes$highschool,positive=1)

```
With an AUC of 0.96 the model is doing great!

```{R}
#Cross validation 

k=5
data <- hate_crimes[sample(nrow(hate_crimes)),] #randomly order rows
folds<-cut(seq(1:nrow(data)),breaks=k,labels=F) #create folds
diags<-NULL
for(i in 1:k){
  ## Create training and test sets
  train<-data[folds!=i,] 
  test<-data[folds==i,]
  truth<-test$highschool
  ## Train model on training set
  fit<-glm(highschool ~ median_house_inc + share_non_white + share_pop_metro + hate_crimes_per_100k_splc, data=train, family="binomial")
  probs<-predict(fit,newdata = test,type="response")
  ## Test model on test set (save all k results)
  diags<-rbind(diags,class_diag(probs,truth, positive=1))
}

summarize_all(diags,mean)
```

The AUC in the CV is 0.80, which is a significant drop. This means that the model does more poorly in the CV and could be a sign of overfitting.

### Non-Parametric Classifier

```{R}
#Fit of non-parametric classifier - k-nearest neighbors

library(caret)
knn_fit <- knn3(factor(highschool==1,levels=c("TRUE","FALSE")) ~ median_house_inc + share_non_white + share_pop_metro + hate_crimes_per_100k_splc, data=hate_crimes, k=5)
y_hat_knn <- predict(knn_fit,hate_crimes)

table(truth= factor(hate_crimes$highschool==1, levels=c("TRUE","FALSE")),
      prediction= factor(y_hat_knn[,1]>.5, levels=c("TRUE","FALSE")))

class_diag(y_hat_knn[,1],hate_crimes$highschool, positive=1)
```

With an AUC of 0.91 the model is doing great!


```{R}
#Cross validation

k=5 #choose number of folds
data<-hate_crimes[sample(nrow(hate_crimes)),] #randomly order rows
folds<-cut(seq(1:nrow(hate_crimes)),breaks=k,labels=F) #create 10 folds
diags<-NULL
for(i in 1:k){
  ## Create training and test sets
  train<-data[folds!=i,] 
  test<-data[folds==i,]
  truth<-test$highschool
  ## Train model on training set
  fit<-knn3(highschool ~ median_house_inc + share_non_white + share_pop_metro + hate_crimes_per_100k_splc,data=train)
  probs<-predict(fit,newdata = test)[,2]
  ## Test model on test set
  diags<-rbind(diags,class_diag(probs,truth, positive=1))
}

summarize_all(diags,mean)
```

The AUC in the CV is 0.87, which means that the model does more poorly in CV. This could be a sign of overfitting.


### Regression/Numeric Prediction

```{R}
#Fit of linear regression model
fit<-lm(mpg~.,data=mtcars) #predict mpg from all other variables
yhat<-predict(fit) #predicted mpg

mean((mtcars$mpg-yhat)^2) #mean squared error (MSE)
```

A MSE of 4.6 is pretty low, but we wan't to compare it to the MSE in the cross validation. 

```{R}
#Cross validation

k=5 #choose number of folds
data<-hate_crimes[sample(nrow(hate_crimes)),] #randomly order rows
folds<-cut(seq(1:nrow(hate_crimes)),breaks=k,labels=F) #create folds
diags<-NULL
for(i in 1:k){
  train<-data[folds!=i,]
  test<-data[folds==i,]
  ## Fit linear regression model to training set
  fit<-lm(highschool ~ median_house_inc + share_non_white + share_pop_metro + hate_crimes_per_100k_splc,data=train)
  ## Get predictions/y-hats on test set (fold i)
  yhat<-predict(fit,newdata=test)
  ## Compute prediction error  (MSE) for fold i
  diags<-mean((test$highschool-yhat)^2) 
}
mean(diags) ## Get average MSE across all folds
```

The Mean Square Error is lower in the CV, which means that the model is predicting good on new data. The model doesn't show signs of overfitting, which also makes sense given that only four predictors are included. 

### Python 

```{R}
library(reticulate)

monthly_income <- hate_crimes$median_house_inc/12
```

```{python}
print(r.monthly_income)
```

For the sharing of objects between R and Python, I created a new variable in an R chunck that gives the median monthly house income in every state and then I used an Python chunck to print out the values of the new variable.